Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-20 infercnv/reports/gains_and_losses.Rmd
add test for markers to gains_and_losses

```{r load_sce, results = 'asis'}
# load gene coords
coords <-
  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))

@@

## `map_metadata_from_infercnv.txt`

@@

@@

Then, we visualise gains and loss proportions for each chromosome on the UMAP. The `Nebulosa` package is used to show the values as densities.

```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = 'hide'}
paste0('chr', 1:22) %>% 
  purrr::map(function(chr) {
    events <-
      paste('proportion', c('dupli', 'loss'), chr, sep = '_')
    sce %>%
      Nebulosa::plot_density(events, size = 0.5) %>%
      print()
    reducedDim(sce, 'UMAP_uwot') %>%
      cbind(sce@colData[, events]) %>%
      tibble::as_tibble() %>%
      tidyr::pivot_longer(3:4) %>%
      dplyr::arrange(value) %>%
      ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) + 
      geom_point(size = 0.1) +
      coord_fixed() +
      theme_classic() +
      scale_color_viridis_c() +
      facet_grid(~ name) %>%
      print()
  })
```

@@

```{r plot_de}
# plot
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == 'chr7', shape = chr == 'chr7')) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = 'dashed', colour = 'purple') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name, ),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey') +
  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
```

@@

>>>>

```{r load_sce, results = 'asis', class.source = 'fold-show'}
# load gene coords
coords <-
  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))

@@

@@

## Cell type compositions

```{r ct_comp, fig.width = 20, fig.height = 7}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    sample = forcats::fct_infreq(sample),
    cell_group = forcats::fct_infreq(cell_group)) %>%
  dplyr::add_count(sample, cell_group, name = 'n_sample_x_cell_group')
p_base <-
  p_dat %>%
  ggplot(aes(fill = cell_group, colour = cell_group)) 
p1 <- 
  p_base +
  geom_point(aes(x = sample, y = cell_group, 
                 alpha = n_sample_x_cell_group, size = n_sample_x_cell_group)) +
  theme_classic() +
  theme(legend.position = 'none') +
  scale_x_discrete(guide = guide_axis(angle = -90))
p2 <- 
  p_base +
  geom_bar(aes(x = sample), position = 'stack') +
  theme_void() +
  theme(legend.position = 'none') 
p3 <-
  p_base +
  geom_bar(aes(y = cell_group), position = 'stack') +
  theme_void() +
  theme(legend.position = 'none')
design <-
'BB#
AAC'
patchwork::wrap_plots(A = p1, B = p2, C = p3, design = design)
```

@@

Then, we visualise gains and loss proportions for each chromosome on the UMAP.

```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = 'hide'}
paste0('chr', 1:22) %>% 
  purrr::map(function(chr) {
    events <-
      paste('proportion', c('dupli', 'loss'), chr, sep = '_')
    reducedDim(sce, 'UMAP_uwot') %>%
      cbind(sce@colData[, events]) %>%
      tibble::as_tibble() %>%
      tidyr::pivot_longer(3:4) %>%
      dplyr::arrange(value) %>%
      ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) + 
      geom_point(size = 0.1) +
      coord_fixed() +
      theme_classic() +
      scale_color_viridis_c() +
      facet_grid(~ name) %>%
      print()
  })
```

@@

```{r plot_de}
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == 'chr7', shape = chr == 'chr7')) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = 'dashed', colour = 'purple') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name, ),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey') +
  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
```

@@

# Cell group markers

We convert the `SingleCellExperiment` object to a `Seurat` object in order to find the top markers of each cell group.

```{r sce_to_seu}
seu <- xfun::cache_rds({
  seu <- Seurat::as.Seurat(sce)
  Seurat::Idents(seu) <- seu$cell_group
  seu <- Seurat::ScaleData(seu)
  seu
}, file = 'seu.rds', rerun = params$rerun)
```

Now, we find all differentially expressed genes for each cell group.

```{r markers}
markers <- xfun::cache_rds({
  markers <-
    Seurat::FindAllMarkers(seu)
  markers
}, file = 'markers.rds', rerun = params$rerun)
```

```{r inspect_markers, fig.height = 10, fig.width = 10}
top2 <-
  markers %>%
  tibble::as_tibble() %>% 
  dplyr::filter(p_val_adj < 0.01, avg_log2FC > 1.5) %>% 
  dplyr::group_by(cluster) %>% 
  dplyr::slice_max(order_by = avg_log2FC, n = 2)
unique(top2$cluster) %>%
  purrr::walk(function(cg) {
    genes <- 
      top2 %>% 
      dplyr::filter(cluster == cg) %>% 
      dplyr::pull(gene)
    list(
      Seurat::VlnPlot(seu, features = genes),
      Seurat::FeaturePlot(seu, features = genes, label = T) + coord_fixed()) %>%
      patchwork::wrap_plots(nrow = 2) +
      patchwork::plot_annotation(
        title = paste0(params$cell_group_name, ': ', cg)) %>%
      print()
  })
```

```{r marker_heatmap, fig.height = 10, fig.width = 10}
top10 <-
  markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 100) %>%
  dplyr::ungroup()
Seurat::DoHeatmap(seu, features = top10$gene) + 
  Seurat::NoLegend()
```
